Graph Topological Aspects of Granger Causal Network Learning

### 2019-07-11 paper/paper.tex
Minor updates, corrects, etc.

%% bibliography stuff -- this needs come before the preamble inclusion
\usepackage[backend=bibtex,sorting=none]{biblatex}
\usepackage{enumitem}
\usepackage{etex,etoolbox}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{fullpage}
\bibliography{\string~/Documents/academics/global_academics/global_bib.bib}
% \usepackage{apxproof}

@@

\todo{Do we have $A_{ij}(z) \ne 0 \implies j \in \anc{i}$?}

@@

  where $(I - \B(z))^{-1} = \A(z)$ due to the uniqueness of
  (\ref{eqn:wold}).  Since $\B(z)$ is stable, that is
  $|\lambda_{\text{max}}(B(z))| < 1$, for every $|z| \le 1$ we have

  \begin{equation}
    \label{eqn:resolvant_inv}
    (I - \B(z))^{-1} = \sum_{k = 0}^\infty \B(z)^k,
  \end{equation}

  where we can apply the Cayley-Hamilton theorem to write the infinite
  sum of \eqref{eqn:resolvant_inv} in terms of \textit{finite} powers
  of $\B$.

  Let $S$ be a $0-1$ matrix containing the sparsity pattern of
  $\B(z)$, from lemma \ref{lem:adj_matrix} $S$ is the transpose of the
  adjacency matrix for $\gcg$ and hence $(S^k)_{ij}$ is non-zero if
  and only if $j \in \gpn{k}{i}$, and therefore $\B(z)^k_{ij} = 0$ if
  $j \not \in \gpn{k}{i}$.  Finally, since
  $\anc{i} = \bigcup_{k = 1}^n\gpn{k}{i}$ we see that $\A_{ij}(z)$ is
  zero if $j \not\in \anc{i}$.

@@

  \begin{align*}
    x_i(t) &= (I - \B(z))^{-1}v(t)\\
    &= \sum_{j = 1}^n \A_{ij}(z) v_j(t)\\
    &= \A_{ii}(z) v_i(t) + \sum_{\substack{j \in \anc{i} \\ j \ne i}} \A_{ij}(z) v_j(t)
  \end{align*}
\end{proof}

@@

This notion is of interest for a variety of reasons.  From a purely
conceptual standpoint, we will see how the notion can in some sense
capture the idea of ``flow of information'' in the underlying graph.
It may also be useful for reasoning about what the conditions under
which \textit{unobserved} components of $x(t)$ may or may not
interfere with inference in the actually observed components.
Finally, motivated from a practical standpoint to analyze causation in
large systems, we will seek to construct practical estimation
procedures based purely on pairwise causality tests since the
computation of such pairwise relations is somewhat easier.
\todo{Reconsider this paragraph}

@@

\begin{remark}
  Note that $H_t^{(i)} \perp \H_t^{(j)}$ states that the
  \textit{present} of $x_i(t)$ is uncorrelated with $x_j(t)$,
  but we need not have $\H_t^{(i)} \perp \H_t^{(j)}$ since
  $i \in \anc{j}$ is not excluded.
\end{remark}

@@

The previous proposition can be strengthened significantly for the
case of Granger-causality.  Note that it is possible to have some
$k \in \anc{i} \cap \anc{j}$ where still $j \npwgc i$, consider for example
the three node graph $k \rightarrow i \rightarrow j$ where
clearly $k \in \anc{i}\cap\anc{j}$ but $j \npwgc i$.  We must
introduce the concept of a \textit{confounding} variable, which
effectively eliminates the possibility presented in this example.

@@

  We notice that for any $u$ having the properties $a, b$ above, we
  must have $C_2(u) = \emptyset$ since if $k \in C_2(u)$ then
  $\exists w \in \anc{i} \cap \anc{k}$ s.t. $i \not \in \anc{k}$ and
  therefore there must be a path $\gcgpath{w}{k} \rightarrow u$ which
  does not contain $i$.

@@

  which follows from Lemma \ref{lem:ancestor_uncorrelated}.
  Finally, for $k \in C_1(j)$ the case $k = i$ is immediate (since the
  error in estimating $\Phi(z)\B_{ii}(z)x_i(t)$ given
  $\H_{t - 1}^{(i)}$ is $0$), so suppose $k \ne i$.  We know from
  above that $k$ inherits the key properties referred to as $a, b$ and
  therefore we can recursively expand $k$ in the same way as in
  equation (\ref{eqn:xj_partition_expansion}).  Continuing this
  recursion for each $k \in C_1(j)$ (where $k \ne i$) must eventually
  terminate since $i \in \anc{k}$.
\end{proof}

@@

\begin{remark}
  As we have seen and as is true in much of statistics, confounding
  nodes pose challenges for Granger-causality.  However, as opposed to
  Pearl's causal calculus \cite{pearl2000art}, pairwise
  Granger-causality does not suffer any difficulty with so-called
  ``colliders'', that is, the topology $i \rightarrow k \leftarrow j$
  will never result in $i \pwgc j$ or $j \pwgc i$.  This is evidently
  an advantage of the \textit{temporal} nature of Granger -- there is
  no backwards causal flow along the edges of $\gcg$.
\end{remark}

@@

\subsection{Persistent Systems}
\label{sec:persistent_systems}
In section \ref{sec:strongly_causal_graphs} we obtained a converse to
part $(a)$ of proposition \ref{prop:ancestor_properties} via the
notion of a strongly causal graph topology.  In this section, we
complete a converse by adding the additional requirement we refer to
as ``persistence''.  This condition requires that each node maintains
some ``memory'' of the past.

@@

  i.e. the ``first'' and ``last'' coefficients of the filter $\B(z)$,
  where $\tau_\infty(\B) \defeq \infty$ if the filter has an infinite
  length.
\end{definition}

\begin{definition}[Persistent]
  We will say that a WSS process $x(t)$ with Granger-causality graph
  $\gcg$ is \textit{persistent} if for every $i \in [n]$, every
  $k \in \anc{i}$ we have $\tau_\infty(\A_{ik}) = \infty$.
\end{definition}

@@

  Moreover, persistence is not the weakest condition necessary for the
  results of this section, the condition
  $\tau_0(\A_{jk}) < \tau_\infty(\A_{ik})$ for each $i, j, k$ such
  that $k \in \anc{i} \cap \anc{j}$ is enough.
  \todo{Write something on the intuition here.}
\end{remark}

@@

  had a solution in $z$ for every $r \ge N$, where $V$ is an
  $n \times n - 1$ full-rank matrix spanning the nullspace of $\beta$,
  and $\bm{\lambda} = (\lambda_1, \ldots, \lambda_n)$. That is,
  iterates of $\Dg(\bm{\lambda})$ applied to $\bm{\lambda}$ would need to remain
  inside $\beta$'s nullspace.  This would imply that

@@

  \todo{This may remain true for any matrix with at least 2 distinct eigenvalues}
  We can conclude by stating that a $\VAR(1)$ system corresponding to
  matrix $B$ is guaranteed to be persistent if $B$ is diagonalizable
  and has more than $1$ distinct eigenvalue.
\end{example}

@@

  \begin{align*}
    \linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}} &= 0\ \forall \text{ strictly causal } F(z)\\
    \iff \inner{F(z)\A_{jk}(z)v_k(t)}{\A_{ik}(z)v_k(t)} &= 0\ \forall \text{ strictly causal } F(z)\\
    \iff \tau_0(\A_{jk}) \ge \tau_\infty(\A_{ik})
  \end{align*}
\end{corollary}
\begin{proof}
  The final equivalence follows immediately from the Lemma.  For the first equivalence we have

@@

\begin{proposition}
  \label{prop:persistence_converse}
  Suppose $\gcg$ is a strongly causal DAG and that $x(t)$ is persistent, then if $k$
  confounds $(i, j)$ we have $i \pwgc j$ and $j \pwgc i$.
\end{proposition}
\begin{proof}
  We will show that $j \pwgc i$, the other being symmetric.  First
  note also that by proposition \ref{prop:sc_graph_common_anc} we
  cannot have $i \in \anc{j}$ or $j \in \anc{i}$ and therefore every
  $k \in \anc{i}\cap\anc{j}$ will be a confounder.

@@

  Evidently, this is $0$ for each $F$ if and only if
  $\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}} = 0$ and
  $\inner{F(z)\A_{jk}(z)v_k(t)}{\A_{ik}(z)v_k(t)} = 0$, which by
  Corollary \ref{cor:time_lag_cancellation} occurs if and only if
  $\tau_0(\A_{jk}) \ge \tau_\infty(\A_{ik})$, which is impossible
  since in general $\tau_0(\A_{jk}) < \infty$ and for a persistent
  system $\tau_\infty(\A_{ik}) = \infty$.
\end{proof}

\subsection{Recovering $\gcg$ via Pairwise Tests}
\label{sec:pairwise_algorithm}
In this section we will show that if the $\gcg$ of a persistent
process is a strongly causal DAG, then it is possible to recover
$\gcg$ via pairwise tests alone.

@@

  The algorithm proceeds first with the parentless nodes $1, 2$ on the
  initial iteration where the edge $(1, 3)$ is added to $E$.  On the
  next iteration, the edges $(3, 4), (2, 4), (3, 5)$ are added, and
  the false edges $(1, 4), (1, 5)$ is excluded due to the paths
  $1 \rightarrow 3 \rightarrow 4$ and $1 \rightarrow 3 \rightarrow 5$
  already present.  Finally, edge $(4, 6)$ is added, and the false
  $(1, 6), (3, 6), (2, 6)$ edges are similarly excluded due to the
  ordering of the inner loop.

@@

If disturbances are Gaussian, we will have the convergence

\begin{equation}
  F_{ij}(p) \overset{\P}{\rightarrow}
  \left\{
    \begin{array}{ll}
      0;\ x_j \npwgc x_i\\
      \infty;\ x_j \pwgc x_i
    \end{array}
  \right. \text{ as } T \rightarrow \infty.  % the '.' after \right is necessary.
\end{equation}

In our finite sample implementation (see Algorithm
\ref{alg:finite_pwgc}) we add edges to $\widehat{\gcg}$ in order of
the decreasing magnitude of $F_{ij}$ instead of proceeding backwards
through $P_{k - r}$ in Algorithm \ref{alg:pwgr}.  This makes greater
use of the information provided by the test statistic $F_{ij}$,
moreover, if $x_i \gc x_j$ and $x_j \gc x_k$, it is expected that
$F_{kj} > F_{ji}$, thereby providing the same effect as proceeding
backwards through $P_{k - r}$.
\subsection{Model Order Selection}
\label{sec:model_order_selection}
There are a variety of methods to choose the filter orders $p_i$ and
$p_j$ (see e.g. \cite{lutkepohl2005new}), but we will focus in
particular on the Bayesian Information Criteria (BIC).  The BIC is
substantially more conservative than the popular alternative Akaiake
Information Criteria (the BIC is also asymptotically consistent), and
since we are searching for \textit{sparse graphs}, we therefore prefer
the BIC, where we seek to \textit{minimize} over $p$:

@@

We carry this out by a simple direct search on each model order
between $0$ and some prescribed $p_\text{max}$.  In practice it is
sufficient to pick $p_\text{max}$ ad-hoc or via some simple heuristic
e.g. plotting the sequence $BIC(p)$ over $p$, though it is not
technically possible to guarantee that the optimal $p$ is chosen
(since there can in general be arbitrarily long lags from one variable
to another).

\subsection{Efficient Model Estimation}
\label{sec:efficient_model_estimation}
In practice, the vast majority of computational effort involved in
implementing our estimation algorithm is spent calculating the error
estimates $\xi_i(p_i)$ and $\xi_{ij}(p_{ij})$.  This requires fitting
a total of $n^2p_{\text{max}}$ autoregressive models, where the most
naive algorithm (e.g. solving a least squares problem for each model)
for this task will consume $O(n^2p_{\text{max}}^4T)$ time, it is
possible to carry out this task in a much more modest
$O(n^2p_{\text{max}}^2 ) + O(n^2p_{\text{max}}T)$ time via the
autocorrelation method which deploys the following autocovariance
estimates in the Yule-Walker equations:\footnote{The particular
  indexing and normalization given in equation
  \ref{eqn:covariance_estimate} is critical to ensure $\widehat{R}$ is
  positive semidefinite.  The estimate can be viewed as calculating
  the covariance sequence of a signal multiplied by a rectangular
  window.}

@@

@@

\subsection{Edge Probabilities and Error Rate Controls}
\label{sec:error_rate_control}
Denote $F_{ij}$ the Granger-causality statistic of equation
\ref{eqn:gc_statistics} with model orders chosen by the methods of
Section \ref{sec:model_order_selection}.  Note that this statistic is
asymptotically $\chi^2(p_{ij})$ distributed, and denote by $G$ the
cumulative distribution function thereof.  We will define the matrix

@@

to be the matrix of pairwise edge inclusion probabilities.  This is
motivated by the hypothesis test where the hypothesis $H_0$ will be
rejected (and thence we will conclude that $x_j \pwgc x_i$) if
$P_{ij} > 1 - \delta$.

The value $\delta$ can be chosen by a variety of methods, in our case
we apply the Benjamini Hochberg criteria \cite{benjamini_hochberg}
\cite{all_of_statistics} to control the false discovery rate of
pairwise edges to a level $\alpha$ (where we generally take
$\alpha = 0.05$).  Applying the extremely simple Bonferroni method is
overly conservative for this application.

@@

After the graph topology $\widehat{\gcg}$ has been estimated via
Algorithm \ref{alg:finite_pwgc}, we refit the entire model with the
specified sparsity pattern directly via ordinary least squares.
Producing graph estimates which are not strongly causal can be
achieved out by performing sequential estimates
$\widehat{x}_1(t), \widehat{x}_2(t), \ldots$ by estimating a strongly
causal graph with the residuals of the previous model as input, and
then refitting on the estimated sparsity pattern.  However we reserve
any analysis of this scheme for future work.

@@

We compare our results against the LASSO.  Motivated by scaling, we
split the squared error term into separate terms, one for each group of
incident edges on a node, and estimate the collection of $n$ incident
filters $\big\{\B_{ij}(z)\big\}_{j = 1}^n$ that minimizes
$\xi_i^{\text{LASSO}}$ in the following:

@@

\begin{figure}
  \caption{PWGC vs adaLASSO (DAG, $q = \frac{2}{n}$)}
  \label{fig:simulation_results_comparison2}
  \includegraphics[width=0.95\textwidth]{new_lasso_comparison_dag_pmax15_simulation.pdf}
  {\scriptsize Figure \ref{fig:simulation_results_comparison2}
    provides results for systems which do not guarantee the
    assumptions of Theorem \ref{thm:scg_recovery}, though the graph
    has a similar level of sparsity.}
\end{figure}

@@

In reference to figure \ref{fig:simulation_results_comparison1} it
should not be overly surprising that our PWGC algorithm performs
better than the LASSO for the case of a strongly causal graph, since
in this case the assumptions which guarantee the correctness of
Algorithm \ref{alg:finite_pwgc} hold.  However, the performance is
still markedly superior in the case of a more general DAG.  We would
conjecture that a DAG having a similar degree of sparsity as an SCG is
likely to be ``close'' to an SCG.  Figure
\ref{fig:simulation_results_dense} illustrates the severe (expected)
degradation in performance as the number of DAG edges increases while
the number of data samples $T$ remains fixed.  For larger values $q$
in this plot, the number of edges in the graph is comparable to the
number of data samples.

In regards scalability, we have observed that performing the $O(n^2)$
pairwise Granger-causality calculations consumes the vast majority
($> 90\%$) of the computation time.  Since this step is trivially
parallelizable, our algorithm also scales well with multiple cores or
multiple machines.  Figure \ref{fig:simulation_results_scaling} is a
demonstration of this scalability, where we are able to estimate
graphs having over $1500$ nodes (over $2.25 \times 10 ^6$ possible
edges) using only $T = 500$ data points, granted, an SCG on this many
nodes is extremely sparse.  However, it is certainly possible to
improve the scalability of the LASSO as applied to time series data by
taking advantage of the block-Toeplitz covariance structure.

\section{Application}
\label{sec:application}
In this section we apply our methods to a real set of EEG data
obtained from the ``EEG Database Data
Set''\footnote{http://archive.ics.uci.edu/ml/datasets/EEG+Database}
\cite{zhang1995event} on the UCI machine learning repository
\cite{uci_mlr}.  This dataset contains 1 second long measurements of
(64 channel) EEG signals from patients who are given visual stimuli.
The subjects in the study are labeled as being either ``control'' or
``alcoholic'', however, we will ignore this label and instead focus on
distinguishing between subjects based on the Granger-causality graphs
inferred from the subject's trials.  Our reasoning is to focus on the
underlying question: ``Does our PWGC algorithm uncover meaningful
Granger-causal connections from EEG data?''.  Focusing only on the
subject label allows us to answer this question without simultaneously
grappeling with the physiological question of whether alcoholic
subjects as a group have discernable differences in their EEG
readings\footnote{Some exploratory analysis actually suggests
  that alcoholic Granger-causality graphs are not substantively
  different}, which is a stronger requirement than simply that there
are meaningful distinctions between the EEG readings of subjects
generally.

@@

The iterated application of PWGC is an example of a heuristic by which
graph estimates that aren't constrained to being strongly causal can
be obtained.  Theoretical analysis of more sophisticated heuristics
than PWGC (Algorithm \ref{alg:finite_pwgc}) is deferred to future work.

@@

Our application Section \ref{sec:application} provides strong evidence
that PWGC is capable of uncovering meaningful features from networks
of time series data by constructing a classifier which accurately
discriminates between subjects in an EEG experiment based purely on
the Granger-causality graph topology inferred by the iterated
application of PWGC.

>>>>

%% bibliography stuff -- this needs come before the preamble inclusion
\usepackage[backend=bibtex,sorting=none]{biblatex}
\usepackage{enumitem}
\usepackage{etex,etoolbox}
\usepackage{hyperref}
\usepackage{fullpage}
\setlength{\marginparwidth}{2cm}
\usepackage{todonotes}
\bibliography{\string~/Documents/academics/global_academics/global_bib.bib}
% \usepackage{apxproof}

@@

@@

  where $(I - \B(z))^{-1} = \A(z)$ due to the uniqueness of
  (\ref{eqn:wold}).  Since $\B(z)$ is stable we have

  \begin{equation}
    \label{eqn:resolvant_inv}
    (I - \B(z))^{-1} = \sum_{k = 0}^\infty \B(z)^k.
  \end{equation}

  Invoking the Cayley-Hamilton theorem allows writing the infinite sum
  of \eqref{eqn:resolvant_inv} in terms of \textit{finite} powers of
  $\B$.

  Let $S$ be a matrix with elements in $\{0, 1\}$ which represents the
  sparsity pattern of $\B(z)$, from lemma \ref{lem:adj_matrix} $S$ is
  the transpose of the adjacency matrix for $\gcg$ and hence
  $(S^k)_{ij}$ is non-zero if and only if $j \in \gpn{k}{i}$, and
  therefore $\B(z)^k_{ij} = 0$ if $j \not \in \gpn{k}{i}$.  Finally,
  since $\anc{i} = \bigcup_{k = 1}^n\gpn{k}{i}$ we see that
  $\A_{ij}(z)$ is zero if $j \not\in \anc{i}$.

@@

  \begin{align*}
    x_i(t) &= [(I - \B(z))^{-1}v(t)]_i\\
    &= \sum_{j = 1}^n \A_{ij}(z) v_j(t)\\
    &= \A_{ii}(z) v_i(t) + \sum_{\substack{j \in \anc{i} \\ j \ne i}} \A_{ij}(z) v_j(t)
  \end{align*}
\end{proof}

@@

This notion is of interest for a variety of reasons.  From a purely
conceptual standpoint, we will see how the notion can in some sense
capture the idea of ``flow of information'' in the underlying graph,
in the sense that if $j \in \anc{i}$ we expect that $j \pwgc i$.  It
may also be useful for reasoning about the conditions under which
\textit{unobserved} components of $x(t)$ may or may not interfere with
inference in the actually observed components.  Finally, motivated
from a practical standpoint to analyze causation in large systems, we
will seek to construct practical estimation procedures based purely on
pairwise causality tests since the computation of such pairwise
relations is somewhat easier.

@@

\begin{remark}
  Note that $H_t^{(i)} \perp \H_t^{(j)}$ states that the \textit{present}
  of $x_i(t)$ is uncorrelated with the whole history of $x_j(t)$, but
  we need not have $\H_t^{(i)} \perp \H_t^{(j)}$ since $i \in \anc{j}$ is
  not excluded.
\end{remark}

@@

The previous proposition can be strengthened significantly; notice
that it is possible to have some $k \in \anc{i} \cap \anc{j}$ where still
$j \npwgc i$, an example is furnished by the three node graph
$k \rightarrow i \rightarrow j$ where clearly
$k \in \anc{i}\cap\anc{j}$ but $j \npwgc i$.  We must introduce the concept
of a \textit{confounding} variable, which effectively eliminates the
possibility presented in this example.

@@

  We notice that for any $u$ having the properties $(a), (b)$ above, we
  must have $C_2(u) = \emptyset$ since if $k \in C_2(u)$ then
  $\exists w \in \anc{i} \cap \anc{k}$ s.t. $i \not \in \anc{k}$ and
  therefore there must be a path $\gcgpath{w}{k} \rightarrow u$ which
  does not contain $i$.

@@

  which follows from Lemma \ref{lem:ancestor_uncorrelated}.  Finally,
  for $k \in C_1(j)$ the case $k = i$ is immediate (since the error in
  estimating $\Phi(z)\B_{ii}(z)x_i(t)$ given $\H_{t - 1}^{(i)}$ is
  $0$), so suppose $k \ne i$.  We know from above that $k$ inherits the
  key properties referred to as $(a)$ and $(b)$ above and therefore we
  can recursively expand $k$ in the same way as in equation
  (\ref{eqn:xj_partition_expansion}).  Continuing this recursion for
  each $k \in C_1(j)$ (where $k \ne i$) must eventually terminate since
  $i \in \anc{k}$.
\end{proof}

@@

\begin{remark}
  As we have seen and as is true in much of statistics, confounding
  nodes pose challenges for Granger-causality.  However, as opposed to
  Pearl's causal calculus \cite{pearl2000art}, pairwise
  Granger-causality does not suffer any difficulty with so-called
  ``colliders'', that is, the topology $i \rightarrow k \leftarrow j$
  will never result in $i \pwgc j$ or $j \pwgc i$.  This is evidently
  an advantage of the \textit{temporal} nature of Granger-causality -- there is
  no backwards causal flow along the edges of $\gcg$.
\end{remark}

@@

\subsection{Persistent Systems}
\label{sec:persistent_systems}
In section \ref{sec:strongly_causal_graphs} we obtained a converse to
part $(a)$ of proposition \ref{prop:ancestor_properties} via the
notion of a strongly causal graph topology.  In this section, we
complete a converse by adding the additional requirement we refer to
as ``persistence''.

@@

  i.e. the ``first'' and ``last'' coefficients of the filter $\B(z)$,
  where $\tau_\infty(\B) \defeq \infty$ if the filter has an infinite
  length, and $\tau_0(\B) \defeq \infty$ if $\B(z) = 0$.
\end{definition}

This interpretation of the following persistence condition is that
each node stores some ``memory'' of the past.

\begin{definition}[Persistent]
  We will say that the process $x(t)$ with Granger-causality graph
  $\gcg$ is \textit{persistent} if for every $i \in [n]$ and every
  $k \in \anc{i}$ we have $\tau_0(\A_{ik}) < \infty$ and $\tau_\infty(\A_{ik}) = \infty$.
\end{definition}

@@

  Moreover, persistence is not the weakest condition necessary for the
  results of this section, the condition
  $\tau_0(\A_{jk}) < \tau_\infty(\A_{ik})$ for each $i, j, k$ such that
  $k \in \anc{i} \cap \anc{j}$ is enough.  The intuition being that nodes
  $i$ and $j$ are not receiving temporally disjoint information from
  $k$.

\end{remark}

@@

  had a solution in $z$ for every $r \ge N$, where $V$ is an
  $n \times n - 1$ full-rank matrix whose columns span the nullspace of $\beta$,
  and $\bm{\lambda} = (\lambda_1, \ldots, \lambda_n)$. That is,
  iterates of $\Dg(\bm{\lambda})$ applied to $\bm{\lambda}$ would need to remain
  inside $\beta$'s nullspace.  This would imply that

@@

  We see from this example that the collection of $\VAR(1)$ systems
  which are not persistent are pathological, in the sense that their
  system matrices have zero measure when viewed as a subset of $\R^{n^2}$.
\end{example}

@@

  \begin{align*}
    \linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}} &= 0\ \forall \text{ strictly causal } F(z)\\
    \iff \inner{F(z)\A_{jk}(z)v_k(t)}{\A_{ik}(z)v_k(t)} &= 0\ \forall \text{ strictly causal } F(z)\\
    \iff \tau_0(\A_{jk}) \ge \tau_\infty(\A_{ik})
  \end{align*}
\end{corollary}
\begin{proof}
  The final equivalence follows immediately from Lemma \ref{lem:time_lag_cancellation}.  For the first equivalence we have

@@

\begin{proposition}
  \label{prop:persistence_converse}
  Suppose $\gcg$ is a strongly causal DAG and that $x(t)$ is
  persistent, then if there exists a $k$ which confounds $(i, j)$ we
  have $i \pwgc j$ and $j \pwgc i$.
\end{proposition}
\begin{proof}
  We will show that $j \pwgc i$, the other being symmetric.  First
  note also that by proposition \ref{prop:sc_graph_common_anc} we
  cannot have $i \in \anc{j}$ or $j \in \anc{i}$ and therefore every
  $k \in \anc{i}\cap\anc{j}$ will be a confounder.

@@

  This is $0$ for every $F$ if and only if for all $F$ and
  $\forall k \in \anc{i} \cap \anc{j}$ we have
  $$\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}} = 0$$ and
  $$\inner{F(z)\A_{jk}(z)v_k(t)}{\A_{ik}(z)v_k(t)} = 0,$$ which by
  Corollary \ref{cor:time_lag_cancellation} occurs if and only if
  $\tau_0(\A_{jk}) \ge \tau_\infty(\A_{ik})$, which is impossible since by persistence
  $\tau_0(\A_{jk}) < \infty$ and  $\tau_\infty(\A_{ik}) = \infty$.
\end{proof}

\subsection{Recovering $\gcg$ via Pairwise Tests}
\label{sec:pairwise_algorithm}
In this section we will show that if the $\gcg$ of a persistent
process is a strongly causal DAG, then it is possible to recover
$\gcg$ via pairwise tests alone.  This is the main conclusion of the
theoretical analysis in this paper.

@@

  The algorithm proceeds first with the parentless nodes $1, 2$ on the
  initial iteration where the edge $(1, 3)$ is added to $E$.  On the
  next iteration, the edges $(3, 4), (2, 4), (3, 5)$ are added, and
  the false edges $(1, 4), (1, 5)$ are excluded due to the paths
  $1 \rightarrow 3 \rightarrow 4$ and $1 \rightarrow 3 \rightarrow 5$
  already being present.  Finally, edge $(4, 6)$ is added, and the false
  $(1, 6), (3, 6), (2, 6)$ edges are similarly excluded due to the
  ordering of the inner loop.

@@

If the estimation procedure is consistent, we will have the following
convergence (in $\P$ or a.s.):

\begin{equation}
  F_{ij}(p) \rightarrow
  \left\{
    \begin{array}{ll}
      0;\ x_j \npwgc x_i\\
      \infty;\ x_j \pwgc x_i
    \end{array}
  \right. \text{ as } T \rightarrow \infty.  % the '.' after \right is necessary.
\end{equation}

In our finite sample implementation (see Algorithm
\ref{alg:finite_pwgc}) we add edges to $\widehat{\gcg}$ in order of
the decreasing magnitude of $F_{ij}$ instead of proceeding backwards
through $P_{k - r}$ in Algorithm \ref{alg:pwgr}.  This makes greater
use of the information provided by the test statistic $F_{ij}$,
moreover, if $x_i \gc x_j$ and $x_j \gc x_k$, it is expected that
$F_{kj} > F_{ji}$, thereby providing the same effect as proceeding
backwards through $P_{k - r}$.
\subsection{Model Order Selection}
\label{sec:model_order_selection}
There are a variety of methods to choose the filter order $p$ (see
e.g. \cite{lutkepohl2005new}), but we will focus in particular on the
Bayesian Information Criteria (BIC).  The BIC is substantially more
conservative than the popular alternative Akaiake Information Criteria
(the BIC is also asymptotically consistent), and since we are
searching for \textit{sparse graphs}, we therefore prefer the BIC,
where we seek to \textit{minimize} over $p$:

@@

We carry this out by a simple direct search on each model order
between $0$ and some prescribed $p_\text{max}$, resulting in a
collection $p_{ij}$ of model order estimates.  In practice, it is
sufficient to pick $p_\text{max}$ ad-hoc or via some simple heuristic
e.g. plotting the sequence $BIC(p)$ over $p$, though it is not
technically possible to guarantee that the optimal $p$ is less than
the chosen $p_\text{max}$ (since there can in general be arbitrarily
long lags from one variable to another).

\subsection{Efficient Model Estimation}
\label{sec:efficient_model_estimation}
In practice, the vast majority of computational effort involved in
implementing our estimation algorithm is spent calculating the error
estimates $\xi_i(p_i)$ and $\xi_{ij}(p_{ij})$.  This requires fitting a
total of $n^2p_{\text{max}}$ autoregressive models, where the most
naive algorithm (e.g. solving a least squares problem for each model)
for this task will consume $O(n^2p_{\text{max}}^4T)$ time, it is
possible to carry out this task in a much more modest
$O(n^2p_{\text{max}}^2 ) + O(n^2p_{\text{max}}T)$ time via the
autocorrelation method
\cite{hayes_statistical_digital_signal_processing} which substitutes
the following autocovariance estimates in the Yule-Walker
equations:\footnote{The particular indexing and normalization given in
  equation \ref{eqn:covariance_estimate} is critical to ensure
  $\widehat{R}$ is positive semidefinite.  The estimate can be viewed
  as calculating the covariance sequence of a signal multiplied by a
  rectangular window.}

@@

It is imperative that the first index in the summation is $\tau + 1$, as
opposed perhaps to $p_\text{max}$ and that the normalization is
$1 / T$, as opposed perhaps to $1 / (T - p_\text{max})$, in order to
guarantee that $\widehat{R}_x(\tau)$ forms a valid (i.e. positive
definite) covariance sequence.  This results in some bias, however the
dramatic computational speedup is worth it for our purposes.

@@

\subsection{Edge Probabilities and Error Rate Controls}
\label{sec:error_rate_control}
Denote $F_{ij}$ the Granger-causality statistic of equation
\ref{eqn:gc_statistics} with model orders chosen by the methods of
Section \ref{sec:model_order_selection}.  We assume that this
statistic is asymptotically $\chi^2(p_{ij})$ distributed (the
disturbances are Gaussian), and denote by $G$ the cumulative
distribution function thereof.  We will define the matrix

@@

to be the matrix of pairwise edge inclusion P-values.  This is
motivated by the hypothesis test where the hypothesis $H_0$ will be
rejected (and thence we will conclude that $x_j \pwgc x_i$) if
$P_{ij} > 1 - \delta$.

The value $\delta$ can be chosen by a variety of methods, in our case
we apply the Benjamini Hochberg criteria \cite{benjamini_hochberg}
\cite{all_of_statistics} to control the false discovery rate of
pairwise edges to a level $\alpha$ (where we generally take
$\alpha = 0.05$).

@@

After the graph topology $\widehat{\gcg}$ has been estimated via
Algorithm \ref{alg:finite_pwgc}, we refit the entire model with the
specified sparsity pattern directly via ordinary least squares.

We note that producing graph estimates which are not strongly causal
can potentially be achieved by performing sequential estimates
$\widehat{x}_1(t), \widehat{x}_2(t), \ldots$ estimating a strongly causal
graph with the residuals of the previous model as input, and then
refitting on the combined sparsity pattern.  We experiment with this
heuristic in our example application of Section \ref{sec:application},
but reserve theoretical analysis for future work.

@@

We compare our results against the adaptive LASSO
\cite{adaptive_lasso_zou2006}, which outperformed substantially both
the LASSO and the grouped LASSO.  Motivated by scaling, we split the
squared error term into separate terms, one for each group of incident
edges on a node, and estimate the collection of $n$ incident filters
$\big\{\B_{ij}(z)\big\}_{j = 1}^n$ that minimizes
$\xi_i^{\text{LASSO}}$ in the following:

@@

\begin{figure}
  \caption{PWGC vs adaLASSO (DAG, $q = \frac{2}{n}$)}
  \label{fig:simulation_results_comparison2}
  \includegraphics[width=0.95\textwidth]{new_lasso_comparison_dag_pmax15_simulation.pdf}

  {\scriptsize Figure \ref{fig:simulation_results_comparison2}
    provides results for systems which do not guarantee the
    assumptions of Theorem \ref{thm:scg_recovery}, though the graph
    has a similar level of sparsity.}
\end{figure}

@@

In reference to figure \ref{fig:simulation_results_comparison1} it
should not be overly surprising that our PWGC algorithm performs
better than the LASSO for the case of a strongly causal graph, since
in this case the assumptions which guarantee the correctness of
Algorithm \ref{alg:finite_pwgc} hold.  However, the performance is
still markedly superior in the case of a more general DAG.  We would
conjecture that a DAG having a similar degree of sparsity as an SCG is
likely to be ``close'' to an SCG.  Figure
\ref{fig:simulation_results_dense} illustrates the severe (expected)
degradation in performance as the number of edges increases while
the number of data samples $T$ remains fixed.  For larger values $q$
in this plot, the number of edges in the graph is comparable to the
number of data samples.

We have also paid close attention to the performance of PWGC in the
very small sample ($T \le 100$) regime (see Figure
\ref{fig:small_T_comparison}), as this is the regime many applications
must contend with.

In regards scalability, we have observed that performing the $O(n^2)$
pairwise Granger-causality calculations consumes the vast majority
($> 90\%$) of the computation time.  Since this step is trivially
parallelizable, our algorithm also scales well with multiple cores or
multiple machines.  Figure \ref{fig:simulation_results_scaling} is a
demonstration of this scalability, where we are able to estimate
graphs having over $1500$ nodes (over $2.25 \times 10 ^6$ possible edges)
using only $T = 500$ data points, granted, an SCG on this many nodes
is extremely sparse.

\section{Application}
\label{sec:application}
In this section we apply our methods to a real set of EEG data
obtained from the ``EEG Database Data
Set''\footnote{http://archive.ics.uci.edu/ml/datasets/EEG+Database}
\cite{zhang1995event} on the UCI machine learning repository
\cite{uci_mlr}.  This dataset contains 1 second long measurements of
(64 channel) EEG signals from patients who are given visual stimuli.
The subjects in the study are labeled as being either ``control'' or
``alcoholic'', however, we will ignore this label and instead focus on
\textit{distinguishing between subjects} based on the
Granger-causality graphs inferred from the subject's trials.  Our
reasoning is to focus on the underlying question: ``Does our PWGC
algorithm uncover meaningful Granger-causal connections from EEG
data?''.  Focusing only on the subject label allows us to answer this
question without simultaneously grappeling with the physiological
question of whether alcoholic subjects as a group have discernable
differences in their EEG readings\footnote{Some exploratory analysis
  actually suggests that alcoholic Granger-causality graphs are not
  substantively different}, which is a stronger requirement than
simply that there are meaningful distinctions between the EEG readings
of subjects generally.

@@

\begin{remark}
  The iterated application of PWGC is an example of a heuristic by
  which graph estimates that aren't constrained to being strongly
  causal can be obtained.  We defer to future work the theoretical
  analysis of more sophisticated heuristics, or algorithms appropriate
  for different topological assumptions.
\end{remark}

@@

Our application in Section \ref{sec:application} provides strong
evidence that PWGC is capable of uncovering meaningful features from
networks of time series data by constructing a classifier which
accurately discriminates between subjects in an EEG experiment based
purely on the Granger-causality graph topology inferred by the
iterated application of PWGC.
